---
title: "Assignment 5: Data Visualization"
author: "Aubrey Knier"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
setwd("/Users/aubreyknier/Desktop/Spring 2022/ENV872_EDA/Environmental_Data_Analytics_2022")
getwd()

library(tidyverse)
library(cowplot)

PeterPaul.chem.nutrients.data <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")

NIMO.litter.data <- 
  read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2 

#Peter Paul Lakes dates
class(PeterPaul.chem.nutrients.data$sampledate)
PeterPaul.chem.nutrients.data$sampledate <- as.Date(
  PeterPaul.chem.nutrients.data$sampledate, format = "%Y-%m-%d")
class(PeterPaul.chem.nutrients.data$sampledate)

#NIWOT ridge dates
class(NIMO.litter.data$collectDate)
NIMO.litter.data$collectDate <- as.Date(
  NIMO.litter.data$collectDate, format = "%Y-%m-%d")
class(NIMO.litter.data$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_linedraw(base_size = 14) +
  theme(axis.text = element_text(color = "grey15", face="bold"), 
        legend.position = "top")
  
theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4

library(viridis)
library(RColorBrewer)
library(colormap)
library(ggplot2)

NTL_LTER_phos_plot <- 
ggplot(PeterPaul.chem.nutrients.data, aes(x=tp_ug, y=po4, color=lakename)) +
  geom_point() + ylim(0,45) + 
    xlab("Total Phosphorous") + ylab("Phosphate") + 
      labs(color="Lake Name") +
        geom_smooth(method="lm", color="black") +
          scale_color_manual(values=c("seagreen", "darkorange2")) +
        facet_wrap(vars(lakename), nrow = 1) 
print(NTL_LTER_phos_plot)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5

PeterPaul.chem.nutrients.data$month <- as.factor(PeterPaul.chem.nutrients.data$month)

#a. temperature
NTL_LTER_temp_boxplot <- 
ggplot(PeterPaul.chem.nutrients.data, aes(x=month, y=temperature_C)) +
  geom_boxplot(aes(color=lakename)) +
    ylab("Temperature (C)") +
      xlab("Month") +
        xlim(c("5","6","7","8","9")) +
          labs(color="Lake name") +
            theme(legend.position = "none") +
        scale_color_manual(values=c("seagreen", "darkorange2")) 
print(NTL_LTER_temp_boxplot)


#b. TP
NTL_LTER_TP_boxplot <-
  ggplot(PeterPaul.chem.nutrients.data, aes(x=month, y=tp_ug)) +
    geom_boxplot(aes(color=lakename)) +
      ylab("TB") +
        xlab("Month") +
           xlim(c("5","6","7","8","9")) +
          scale_color_manual(values=c("seagreen", "darkorange2")) + theme(legend.position = "none")
print(NTL_LTER_TP_boxplot)

#c. TN
NTL_LTER_TN_boxplot <-
  ggplot(PeterPaul.chem.nutrients.data, aes(x=month, y=tn_ug)) +
    geom_boxplot(aes(color=lakename)) +
      ylab("TN") +
        xlab("Month") +
          xlim(c("5","6","7","8","9")) +
          scale_color_manual(values=c("seagreen", "darkorange2")) + theme(legend.position = "none")
print(NTL_LTER_TN_boxplot)


legend_NTL_LTER <- get_legend(
  NTL_LTER_temp_boxplot + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)

plot_grid(NTL_LTER_temp_boxplot, NTL_LTER_TP_boxplot, NTL_LTER_TN_boxplot, legend_NTL_LTER, nrow = 2, align = 'h', rel_heights = c(1.25, 1.25))

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6

#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
